### 2022-06-08 paper/paper.tex
trainer: add initial trainer

% Code Formatting
\usepackage{minted}
\usemintedstyle{pastie}

\def\code#1{\texttt{#1}}

@@

@@

\addto\captionsamerican{% Replace "english" with the language you use
	\renewcommand{\contentsname}%
	{Table of Contents}%
}

@@

	\hrulefill{}
	\par\vspace{0.25cm}
	A Special Project Proposal Presented to the\\
	Faculty of the Department of Computer Science,\\
	University of the Philippines Cebu

@@

	\textbf{\uppercase{Development of a finger-key identification module\\for a touch typing trainer}}

@@

\noindent
This thesis entitled ``\textbf{\uppercase{Development of a finger-key
		identification \\module for a touch typing trainer}}'' prepared and
submitted by \\\uppercase{Oscar Vian L. Valles} is approved by:

@@

\begin{center}
	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Aileen Joan O. Vicente\\
		Panel Member

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date
	\end{minipage}
	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Aileen Joan O. Vicente\\
		Panel Member

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date
	\end{minipage}

@@

	\line(1,0){0.405\textwidth}\\
	Prof. Aileen Joan O. Vicente\\
	Panel Chair

	\par\vspace{0.5cm}
	\line(1,0){0.405\textwidth}\\
	Date

	\begin{minipage}{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Aileen Joan O. Vicente\\
		Chair, Department of Computer Science

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date
	\end{minipage}
	\begin{minipage}{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Nelia S. Ereno\\
		Dean, College of Science

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date
	\end{minipage}
\end{center}

@@

@@

\section{Background of the Study}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests
list out words that are then typed out. The entered keys are then compared to
check if the user has typed the expected letter. At the end of the test, the
time taken is calculated, and certain metrics are given. These metrics include
\ac{wpm} and Accuracy \parencite{bartnik2021}.

However, this method of examination leaves out a crucial part of typing —
ergonomics. Ergonomic typing prevents a lot of health issues in the future like
repetitive strain injury or carpal tunnel. One important factor that affects
ergonomics is the typing procedure and posture. This means the proper placement of
the wrist, hands, and hitting the keys using the right finger that is assigned
to the key.

Correct finger placement is usually taught at the beginning using a diagram,
with each key being associated with a specific finger. For instance, the letter
Q in a QWERTY layout should be hit using the fifth digit, or the little finger,
of the left hand, and this is shown by coloring the fifth digit and the key Q
with the same color or by placing the letters directly on the fingers
\parencite{dobson2009touch}.

Incorrect finger placement may cause these hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \parencite{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions may cause injuries such as \ac{cts}
\parencite{toosi2015}

In addition, this type of typing is frequently taught in the beginner level
\parencite{donica2018}. This means that there is a need to weed out bad habits
that may develop, like using the index finger for pressing the spacebar or
backspace. However, it is impractical for an educator to check each student if
they are not performing these movements as these may only show for a small
period which may not be caught in time.

@@

\section{Research Questions}
\begin{enumerate}
	\item What algorithm or technique using computer vision is capable of
	      finger-key identification.
	\item What setup and configuration of a single optical camera can be used to
	      capture images during keyboard typing for time finger-key identification
	      using the algorithm identified in 1?
	\item How to design and develop a finger-key identification module using
	      the algorithm in 1 and camera setup and configuration in 2?
	\item How accurate is the module developed in 3 in finger-key identification?
	\item Is the module developed in 3 fast enough to run in real-time?
	\item What new keyboard typing metric to develop that considers both the key and
	      finger used to press as parameters?
\end{enumerate}

@@

\subsection{Specific Objectives}
\begin{enumerate}
	\item Develop an algorithm or technique using computer vision that is capable
	      of finger-key identification.
	\item Identify the setup and configuration of a single optical camera that can
	      be used to capture images during keyboard typing for finger-key
	      identification using the algorithm identified in 1
	\item Design and develop a module using the algorithm in 1 and camera setup
	      and configuration in 2
	\item Determine the accuracy of the module developed in 3 in finger-key
	      identification
	\item Determine if the module developed in 3 is fast-enough to run in
	      real-time
	\item Develop a new keyboard typing metric that uses both the key and
	      finger used to press as parameters
\end{enumerate}

\section{Scope and Limitation}
This research will focus on typing on a 60\% keyboard. Figure~\ref{fig:60}
illustrates this type of keyboard. This type of keyboard only has the
alphanumeric part of the keyboard. This limits the number of keys to be checked
and the expected movement of the hand. Furthermore, the keycaps will also be of
a light color, while the surface that the keyboard rests upon will be of a dark
color.

In addition, the keyboard layout will be \ac{ansi}. This layout is described by
the American National Standards Institute \parencite{ansi}. This is the most
common layout in the United States. However, it is also used in numerous
English-speaking countries such as the Philippines, Malaysia, and India
\parencite{apple-layout}.

The setup that will be captured will be simplified and fixed for the development
and testing of the finger-key identification module. This includes the
keyboard, surface it is placed on, and lighting of the entire area. Furthermore,
the setup will not be tested by other respondents, but by the researchers only
--- as the focus is on the development and testing of the finger-key
identification module and not on the touch typing trainer and its effects on
typing.

The program will expect the that user has all ten digits and has no hand,
finger, or wrist deformities. In addition, only the placement of the hands,
fingers, and wrists will be taken into account when determining if the
ergonomics of the user while typing is healthy. The program will not check
seating position, angle of elbows, and other metrics for an ergonomic typing
posture while typing.

Capturing of the video to be analyzed by the program would be limited to a
single 720p webcam that is capturing in 30 frames per second. The camera will be
pointed downwards facing the keyboard and the hand.

@@

Keyboard typing is the process of using a keyboard to input characters in a
system. In the context of this paper, keyboard typing will refer to the act of
using a physical keyboard to input characters in a computer system.

@@

\citeauthor{ansi} is the standard that first defined the \ac{ansi} layout.
Figure~\ref{fig:ansi} illustrates what the \ac{ansi} layout looks like. This layout
is also used by countries other than America. Examples of countries that use
this layout as its standard is the Philippines, China, and Korea
\parencite{apple-layout}. However, these countries also opt to modify the layout
by adding extra layers to accommodate other character sets.

\citeauthor{iso} is the standard series that defines a framework that is used to
create other layouts. Layouts created from this standard are colloquially called
\ac{iso} Layouts. Countries around the world use this framework to create layouts
that fit the characters in their language. Examples of countries that use this
framework to create their own layout are France, Greece, Canada, and Sweden
\parencite{apple-layout}.

@@

There are other layouts available, however, they are not as common as the two
previously mentioned layouts. Examples of this include \ac{jis}. Other esoteric
layouts, like Tsangan or split-backspace, also exist. These layouts modify the
\ac{iso} and \ac{ansi} standards by adding or removing certain keys to fit the
character set of a language, or for additional keys. Other layouts are also
exactly the same as \ac{ansi} or \ac{iso}, however, these layouts change the
arrangement of the alphabet within the keyboard.

Despite the ubiquity of these common layouts, studies have shown that these
layouts are not ergonomic. The main issue with these layouts is the random
configurations of the letters. The randomness of the layout necessitates
memorization of the layout which reduces the ease of learning, reduces performance
in typing by reducing speed, and increases of typing errors
\parencite{ciobanu2015}.

@@

There is only one common keyboard form factor used worldwide: the full-size
keyboard. This keyboard contains all the keys specified in the keyboard
layout. This includes the alphanumeric keys, the function keys, the navigation
cluster, and the numpad.

@@

\subsubsection{Speed}
Speed, also called as entry rate by \citeauthor{arif2009}, measures the number
of characters entered in a specific time frame. The most common metric that
measures speed is \ac{wpm}. \ac{wpm} as defined by \citeauthor{arif2009} is:

@@

\subsubsection{Average Typing Speed}

Average typing speed using a keyboard with QWERTY layout, based on test data
from \cite{keybr}, is $\approx37.5$ \ac{wpm}. This average is taken from all of
the test results they have gathered from all of their users.
\cite{arif2009} also performed a survey with 12 participants using
multiple typing methods. They got an average \ac{wpm} of 75.85 with a standard
deviation of 15.61 when their participants used a physical QWERTY layout. This
average \ac{wpm} was also the highest among the other typing methods the
researchers studied.

@@

These studies start teaching kids in the kindergarten level and the studies try
to optimize the teaching methods to improve the speed and accuracy of typing of
the learners. By starting to teach touch typing to students early, these
students will develop the potential for higher-level keyboard typing
\parencite{donica2018}.

@@

Through this process, the learner goes through the three stages of Motor
Learning Theory. The student undergoes the cognitive stage where they try to
understand and create strategies to accomplish the given task. Then the
associative stage follows where the strategies and skills learned from the
previous stage are refined. At this stage, the learners are expected to rely less
on visuals to locate the keys and more on kinesthesis. By the final stage, the
autonomous stage, the learner does not rely on visuals at all and focuses on
using kinesthetic feedback to find the keys. By this point, the learner has
progressed from using Hunt and Peck, to becoming proficient in touch typing.
\parencite{donica2018}

@@

Keybr is similar to Monkeytype and Typerace, in that they also have the users
type a predetermined phrase, quote, or random words. However, this application
has more guidance compared to the two. Keybr uses statistics to create typing
lessons that are appropriate to the current typing proficiency of the learner.
The words selected are random at first, and the skill level of the learner is
determined by the performance of the user with these words and characters. The
information gathered is then used to generate new words for the next iteration.
As an example, if a learner has difficulty in typing the letter q, the next
iterations will have a lot of words that contain the letter q.

@@

There have been a lot of studies that show the effect of keyboard typing, and
its associated movements (or lack thereof), has an effect on the human body.
These studies have shown that keyboard typing has an effect on our neck, shoulder,
upper limb, wrist, arms, and fingers \parencite{szeto2005, baker2007digit}

@@

\subsection{Finger and Wrist Kinematics}
The way people move their hands, wrists, and fingers differ between each person.
This can be attributed to the different typing styles each person has. One key
difference between people is the angle of the fifth digit.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\cite{marklin1999} as cited in
~\cite{baker2007})

\section{Finger and Hand Tracking}
Finger and Hand tracking is a method of tracking fingers and hands in 3D space
using motion capture systems or computer vision. This technique allows computers
to perform actions and analyzes on the motions and positions of these body
parts.

@@

But this method is cost-prohibitive, and cannot handle occlusions well.
\citeauthor{alexanderson2016} present a method for an optical marker-based
motion capture system that can predictably recover from self-occlusion and has a
better performance compared to previously used algorithms, however, the issue of
cost and self-occlusion still persists.

@@

At its core, Computer Vision aims to perform tasks that the human visual system
can do \parencite{cern}. This includes object classification, tracking, and
gesture recognition, and face recognition. At the present, most computer vision
systems utilize deep learning algorithms, and convolutional networks to gather
information from an image, or a set of images. One such example of a
convolutional network used in computer vision is Inception by
\citeauthor{szegedy2015} which proposes a convolutional neural network
architecture for object classification and detection.

@@

This framework allows for performant object detection, hand and finger tracking,
human pose detection. The framework also allows for combining multiple features,
by adding them to the graph as calculators. MediaPipe has C++, Python, JS, and
Coral interfaces. It also supports Android and iOS devices
\parencite{mediapipe}.

@@

A toolbox offered by MATLAB is the Computer Vision Toolbox that contains
algorithms, and functions for use in the development of computer vision, 3D
vision, and video processing systems. By using the available algorithms in the
toolbox, such as YOLOv2, and ACF, hand detection and gesture recognition is made
possible in the platform \parencite{matlab}.

@@

\cite{dorf2001} presents a use case for finger tracking in augmented
environments. In the paper, interaction in a virtual environment through the use
of gestures. The tracking system uses an optical marker-based motion capture
system where the user wears a glove with retroreflective markers.

@@

\section{Summary of the Research Gap}
While there are a lot of applications and curriculum aimed at teaching touch
typing, there is no automated system available that detects if a person uses the
correct finger to press a key.

@@

The experimental setup and configuration was composed of three elements: the
camera, the keyboard, and the environment. Figure~\ref{fig:metho-setup} is an
image of the experimental setup

@@

\subsection{Camera}
The setup used a single monocular camera positioned in a top-down view. The
camera captured the entirety of the keyboard and the movement of the ten
fingers. To do so, it was mounted on top of the monitor with the camera pointed
down towards the table.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.4\textwidth]{webcam.png}
	\caption{Logitech C920. Reprinted from \fullcite{logitech}}
	\label{fig:metho-setup-cam}
	\centering
\end{figure}

@@

The keyboard used was a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This limited the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice also
reduced the area that the camera captured, as this keyboard type is considerably
smaller compared to a full-size keyboard. The keyboard also had its keycaps and
case in a light color that contrasted the dark surface it was placed on. It also
had a dark USB-C cable to blend with the dark surface. These color coordination
steps improved initial keyboard detection. In addition, the keyboard layout was
ANSI, due to the availability and widespread adoption of the layout in the
Philippines.

@@

In addition, the surface where the keyboard was placed on was solid black
without any variation of color. This also improved initial keyboard detection.

@@

A computer vision algorithm was created as a starting point
mapping the keys of the keyboard within a video. A rough flowchart of the algorithm
is shown in Figure~\ref{fig:metho-algo-key-flow}.

@@

\paragraph{Convert to Grayscale}
The input frame from the camera was converted to a 256 level grayscale image
using \texttt{cvtColor} of OpenCV \parencite{opencv-cvtColor}. This step was
performed because future steps of the algorithm did not require color values to
work. In addition, this step optimized the algorithm as the number of dimensions
analyzed was reduced.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{grayscale.png}
	\caption{Camera capture converted to grayscale}
	\centering
\end{figure}

\paragraph{Denoising}
The algorithm denoised the grayscale image using a bilateral filter as
implemented by OpenCV \parencite{opencv-bilateral-filter}. This filter takes the
range of the image into account, rather than just the domain. This resulted in
an image that is smoothed while preserving its edges
\parencite{bilateral-filter}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{denoised.png}
	\caption{Smoothed image with intact edges}
	\centering
\end{figure}

\paragraph{Edge Detection}
The algorithm used a Sobel filter for edge detection. This filter uses a
3$\times$3 kernel that is convolved twice. Once horizontally, and another
vertically to produce a grayscale image of the outlines within the frame. The
kernels used by the Sobel filter \parencite{sobel2014} is shown in
Figure~\ref{fig:metho-algo-key-sobel}.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{sobel.png}
	\caption{Output of the Sobel filter}
	\centering
\end{figure}

\paragraph{Thresholding}
The output of the Sobel filter is a grayscale image of 255 values, with each
gray pixel indicating edges within the image. There is a need to reduce the
range of these values to improve the ability of the algorithm to find the
contours of these edges. To do so, the algorithm utilized Otsu's algorithm to
perform automated thresholding. Otsu's algorithm determines a single threshold
that is most optimal for the image \parencite{otsu}. This outputs a
black-and-white image, with only two values.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{threshed.png}
	\caption{Threshed image of the edges}
	\centering
\end{figure}

\paragraph{Find Contours}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function
\texttt{findContours}, with the \texttt{CHAIN\_APPROX\_SIMPLE} contour
approximation method, was used to find the contours of the outlines of the
object found using the previous step. This approximation method removed
redundant points and returned the least amount of points that describes the
shape. This OpenCV function implements the algorithm of~\cite{contours} in their
paper \citetitle{contours}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{find-contours.png}
	\caption{All the contours found in the image}
	\centering
\end{figure}

@@

The contour points that described the contour were then sorted clockwise. This
was a necessity since the algorithm of~\cite{contours} does not guarantee that
the largest contour's points were returned in a clockwise arrangement, which is
a requirement for the next step

>>>>

@@

\DeclareAcronym{cps}{
	short=CPS,
	long=Characters per Second
}

@@

\addto\captionsamerican{
	\renewcommand{\contentsname}
	{Table of Contents}
}

% Subfigure
\usepackage{subcaption}

@@

	\hrulefill{}
	\par\vspace{0.25cm}
	A Special Project Presented to the\\
	Faculty of the Department of Computer Science,\\
	University of the Philippines Cebu

@@

	\textbf{\uppercase{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}}

@@

\begin{center}
	\uppercase{\textbf{APPROVAL SHEET}}
\end{center}

\par\vspace{0.5cm}
\noindent
The faculty of the University of the Philippines Cebu, College of Science,
Department of Computer Science approves this Special Problem entitled:
\par\vspace{0.5cm}

\begin{center}
	\textbf{Development of a Finger-Key Identification Module\\for a Touch Typing Trainer}
	\par\vspace{0.5cm}
	by
	Oscar Vian L. Valles

@@

	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Dhong Fhel K. Gom-os\\
		Special Problem Adviser

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date Signed
	\end{minipage}
	\begin{minipage}[t]{0.45\textwidth}
		\centering
		\line(1,0){0.9\textwidth}\\
		Prof. Aileen Joan O. Vicente\\
		Chair, Department of Computer Science

		\par\vspace{0.5cm}
		\line(1,0){0.9\textwidth}\\
		Date Signed
	\end{minipage}

@@

	\line(1,0){0.405\textwidth}\\
	Prof. Nelia S. Ereno\\
	Dean, College of Science

	\par\vspace{0.5cm}
	\line(1,0){0.405\textwidth}\\
	Date Signed

	\par\vspace{2cm}
\end{center}

@@

Typing tests lists out words for the user to repeat. These tests measure metrics
that quantify the user's ability to type. However, conventional typing metrics
do not measure correct finger placement. This aspect of typing may affect the
user's health. It is also a crucial aspect of keyboard typing education. As
such, a method to identify which finger is used to press which key is
beneficial. This paper introduces a technique to do so by developing a
finger-key identification module that utilizes computer vision algorithms to
detect the keys in a keyboard, and a ready-made machine learning solution to
track fingers while typing. This module was successful in finger-key
identification with an accuracy of 99.47\% in a dataset of 937 keypresses. The
module also had an average finger-key identification time of 0.083 seconds,
which allows the module to perform real-time finger-key identification for
typing speeds up to 143.030 Words per Minute which is well beyond the median
typing speed of the population. Finally, two metrics for correct finger
placement were defined. Finger Placement Accuracy which computes the number of
keys pressed with the correct finger in a typing test sequence. Historical
Finger Placement Accuracy which evaluates the user's ability to press a certain
key with the correct finger over multiple typing test sequences.

@@

\section{Background of the Study}
There are a lot of educational typing tests available that help people learn
touch typing, including Monkeytype, TypeRacer, and Keybr. These typing tests
list out words that are then typed out. The entered keys are then compared to
check if the user has typed the expected letter. At the end of the test, the
time taken is calculated, and certain metrics are given. These metrics include
\ac{wpm} and Accuracy \parencite{arif2009}. However, this method of examination
leaves out a crucial part of typing --— correct finger placement.

Incorrect finger placement may affect hand and finger health. Incorrect
placement of the fingers may cause these hand and wrist positions: ulnar
deviation, forearm pronation, and wrist extension \parencite{serina1999}. These
three are hand and wrist positions that are common in all activities, however,
prolonged periods in these positions, such as in typing, may cause injuries such
as \ac{cts} \parencite{toosi2015}.

Correct finger placement is usually taught at the beginning using a diagram,
with each key being associated with a specific finger. For instance, the letter
Q in a QWERTY layout should be pressed using the fifth digit, or the little
finger, of the left hand, and this is shown on instruction manuals by coloring
the fifth digit and the key Q with the same color or by placing the letters
directly on the fingers \parencite{dobson2009touch}.

Furthermore, Touch typing using correct finger placement is frequently taught in
the beginner level \parencite{donica2018}. This means that there is a need to
weed out bad habits that may develop, like using the index finger for pressing
the spacebar or backspace. However, it is impractical for an educator to check
each student if they are not performing these movements as these may only show
for a small period and may not be caught in time.

@@

\section{Research Questions}
\begin{enumerate}
	\item What algorithm or technique using computer vision is capable of
	      finger-key identification.
	\item What is the setup and configuration of a single optical camera can be
	      used to capture images during keyboard typing for real-time finger-key
	      identification using the algorithm identified in 1?
	\item How to design and develop a finger-key identification module using
	      the algorithm in 1 and camera setup and configuration in 2?
	\item How accurate is the module developed in 3 in finger-key identification?
	\item Is the module developed in 3 fast enough to run in real-time?
	\item What new keyboard typing metric to develop that considers both the key and
	      finger used to press as parameters?
\end{enumerate}

@@

\subsection{Specific Objectives}
\begin{enumerate}
	\item Develop an algorithm or technique using computer vision capable of
	      finger-key identification.
	\item Identify the setup and configuration of a single optical camera that can
	      be used to capture images during keyboard typing for finger-key
	      identification using the algorithm identified in 1
	\item Design and develop a module using the algorithm in 1 and camera setup
	      and configuration in 2
	\item Determine the accuracy of the module developed in 3 in finger-key
	      identification
	\item Determine if the module developed in 3 is fast enough to run in
	      real-time
	\item Develop a new keyboard typing metric that uses both the key and
	      finger used to press as parameters
\end{enumerate}

\section{Scope and Limitation}
The experimental setup was simplified and tightly controlled. It was not
representative of real-world conditions. The colors of the keyboard, cables, and
surface were all controlled. So were the lighting and camera used.

The supported keyboard of the module was limited to one specific type of
keyboard. A 60\% Keyboard in a slightly modified \ac{ansi} layout with
light-colored keys as shown in Figure~\ref{fig:60}. This type of keyboard only
has the alphanumeric portion of a full-size keyboard. This limited the number of
keys to be checked and reduced the overall movement of the hand during typing.
The keyboard layout was a slightly modified \ac{ansi} layout. The majority of
the layout still followed the base \ac{ansi} layout, with the alphanumeric
portion remaining exactly the same.

The module's effect on typing was not tested. This includes the module's effect
on typing education and typing health, as the focus of the research was on the
development and testing of the finger-key identification module only.

In addition, the module expected the user to have no deformities in their hands,
and they were capable of typing on a keyboard with difficulties.

@@

Keyboard typing is the process of using a keyboard to input characters in a
system. In the context of this paper, keyboard typing will refer to the act of
using a physical keyboard to input characters into a computer system.

@@

\citeauthor{ansi} is the standard that first defined the \ac{ansi} layout.
Figure~\ref{fig:ansi} illustrates what the \ac{ansi} layout looks like. This
layout is also used by countries other than America. Examples of countries that
use this layout as its standard are the Philippines, China, and Korea
\parencite{apple-layout}. However, these countries also opt to modify the layout
by adding extra layers to accommodate other character sets.

\citeauthor{iso} is the standard series that defines a framework that is used to
create other layouts. Layouts created from this standard are colloquially called
\ac{iso} Layouts. Countries around the world use this framework to create
layouts that fit the characters in their language. Examples of countries that
use this framework to create their layout are France, Greece, Canada, and Sweden
\parencite{apple-layout}.

@@

There are other layouts available, however, they are not as common as the two
previously mentioned layouts. Examples of this include \ac{jis}. Other esoteric
layouts, like Tsangan or split-backspace, also exist. These layouts modify the
\ac{iso} and \ac{ansi} standards by adding or removing certain keys to fit the
character set of a language, or for additional keys. Other layouts are also the
same as \ac{ansi} or \ac{iso}, however, these layouts change the arrangement of
the alphabet within the keyboard.

Despite the ubiquity of these common layouts, studies have shown that these
layouts are not ergonomic. The main issue with these layouts is the random
configurations of the letters. The randomness of the layout necessitates
memorization of the layout which reduces the ease of learning, reduces performance
in typing by reducing speed, and increases typing errors
\parencite{ciobanu2015}.

@@

There is only one common keyboard form factor used worldwide: the full-size
keyboard. This keyboard contains all the keys specified in the keyboard
layout. This includes the alphanumeric keys, the function keys, the navigation
cluster, and the Numpad.

@@

\subsubsection{Speed}
Speed, also called as entry rate by \citeauthor{arif2009}, measures the number
of characters entered within a specific time frame. The most common metric that
measures speed is \ac{wpm}. \ac{wpm} as defined by \citeauthor{arif2009} is:

@@

\subsection{Typing Speed Statistics}

Median typing speed using a keyboard with QWERTY layout, based on test data
from \cite{keybr}, is $\approx37.5$ \ac{wpm}. This median is taken from all the test results they have gathered from all of their users.

\cite{monkey-stats} released a dataset showing the speed of 12678 members of the
Monkeytype community with the speed role. The dataset shows the fastest time
that these members have achieved on 60-second tests. The median of the dataset
lies in the 100--109 \ac{wpm} range, with 52.74\% of the community having top
speeds slower than 100 \ac{wpm}.

\cite{arif2009} also performed a survey with 12 participants using
multiple typing methods. They got an average \ac{wpm} of 75.85 with a standard
deviation of 15.61 when their participants used a physical QWERTY layout. This
average \ac{wpm} was also the highest among the other typing methods the
researchers studied.

@@

These studies start teaching kids at the kindergarten level and the studies try
to optimize the teaching methods to improve the speed and accuracy of typing of
the learners. By starting to teach touch typing to students early, these
students will develop the potential for higher-level keyboard typing
\parencite{donica2018}.

@@

Through this process, the learner goes through the three stages of Motor
Learning Theory. The student undergoes the cognitive stage where they try to
understand and create strategies to accomplish the given task. Then the
associative stage follows where the strategies and skills learned from the
previous stage are refined. At this stage, the learners are expected to rely less
on visuals to locate the keys and more on kinesthesis. In the final stage, the
autonomous stage, the learner does not rely on visuals at all and focuses on
using kinesthetic feedback to find the keys. By this point, the learner has
progressed from using Hunt and Peck, to becoming proficient in touch typing.
\parencite{donica2018}

@@

Keybr is similar to Monkeytype and Typerace, in that they also have the users
type a predetermined phrase, quote, or random words. However, this application
has more guidance compared to the two. Keybr uses statistics to create typing
lessons that are appropriate to the current typing proficiency of the learner.
The words selected are random at first, and the skill level of the learner is
determined by the performance of the user with these words and characters. The
information gathered is then used to generate new words for the next iteration.
As an example, if a learner has difficulty typing the letter q, the next
iterations will have a lot of words that contain the letter q.

@@

There have been a lot of studies that show the effect of keyboard typing, and
its associated movements (or lack thereof), affects the human body. These
studies have shown that keyboard typing affects our neck, shoulder,
upper limb, wrist, arms, and fingers \parencite{szeto2005, baker2007digit}

@@

\subsection{Finger and Wrist Kinematics}
The way people move their hands, wrists, and fingers differs between each
person. This can be attributed to the different typing styles each person has.
One key difference between people is the angle of the fifth digit.

@@

For those that reach their keys by stretching their fingers, there is an
increased probability that the wrists and fingers adapt to non-neutral postures.
These include wrist extension, ulnar deviation, and pronation, which may cause
musculoskeletal disorders of the upper extremity (\cite{marklin1999} as cited in
~\cite{baker2007})

\section{Finger and Hand Tracking}
Finger and Hand tracking is a method of tracking fingers and hands in 3D space
using motion capture systems or computer vision. This technique allows computers
to perform actions and analyzes the motions and positions of these body parts.

@@

But this method is cost-prohibitive, and cannot handle occlusions well.
\citeauthor{alexanderson2016} present a method for an optical marker-based
motion capture system that can predictably recover from self-occlusion and has a
better performance compared to previously used algorithms, however, the issue of
cost and self-occlusion persists.

@@

At its core, Computer Vision aims to perform tasks that the human visual system
can do \parencite{cern}. This includes object classification, tracking, gesture
recognition, and face recognition. At the present, most computer vision systems
utilize deep learning algorithms, and convolutional networks to gather
information from an image, or a set of images. One such example of a
convolutional network used in computer vision is Inception by
\citeauthor{szegedy2015} which proposes a convolutional neural network
architecture for object classification and detection.

@@

This framework allows for performant object detection, hand and finger tracking,
and human pose detection. The framework also allows for combining multiple
features, by adding them to the graph as calculators. MediaPipe has C++, Python,
JS, and Coral interfaces. It also supports Android and iOS devices
\parencite{mediapipe}.

@@

A toolbox offered by MATLAB is the Computer Vision Toolbox which contains
algorithms, and functions for use in the development of computer vision, 3D
vision, and video processing systems. By using the available algorithms in the
toolbox, such as YOLOv2, and ACF, hand detection and gesture recognition is made
possible in the platform \parencite{matlab}.

@@

\cite{dorf2001} present a use case for finger tracking in augmented
environments. In the paper, interaction in a virtual environment through the use
of gestures. The tracking system uses an optical marker-based motion capture
system where the user wears a glove with retroreflective markers.

@@

\section{Summary of the Research Gap}
While there are a lot of applications and curricula aimed at teaching touch
typing, there is no automated system available that detects if a person uses the
correct finger to press a key.

@@

The experimental setup and configuration were composed of three elements: the
camera, the keyboard, and the environment. Figure~\ref{fig:metho-setup} is an
image of the experimental setup

@@

\subsection{Camera}
The setup used a single monocular camera positioned in a top-down view. The
camera captured the entirety of the keyboard and the movement of the ten
fingers. To do so, it was mounted on top of the monitor with the camera pointed
down towards the table. Figure~\ref{fig:metho-setup-camera-placement}
illustrates how the camera was mounted over the monitor and
Figure~\ref{fig:metho-setup-keeb} is a picture taken with the camera at this
position.

@@

@@

The keyboard used was a 60\% keyboard as shown in
Figure~\ref{fig:metho-setup-keeb}. This limited the necessary mapping for the
algorithm to the alphanumeric portion of the keyboard. This keyboard choice also
reduced the area that the camera captured, as this keyboard type is considerably
smaller compared to a full-size keyboard. The keyboard also had its keycaps and
case in a light color that contrasted the dark surface it was placed on. It also
had a dark USB-C cable to blend with the dark surface. These color coordination
steps improved initial keyboard detection.

In addition, the keyboard layout was a modified \ac{ansi} layout. The majority
of the keys remained the same, however some keys were changed: The pipe key was
replaced by the backspace key, the backspace key was replaced by the pipe key
and a delete key, and the bottom row has a 7u length spacebar. Overall, these
are minor changes that does not affect the alphanumeric keys that are
predominantly used in typing.

@@

In addition, the surface where the keyboard was placed was solid black
without any variation of color. This also improved initial keyboard detection.

@@

A computer vision algorithm was created as a starting point for mapping the keys
of the keyboard within a video. A rough flowchart of the algorithm is shown in
Figure~\ref{fig:metho-algo-key-flow}.

@@

\paragraph{Convert to Grayscale}
The input frame from the camera was converted to a 256 level grayscale image
using \texttt{cvtColor} of OpenCV \parencite{opencv-cvtColor}. This step was
performed because future steps of the algorithm did not require color values to
work. In addition, this step optimized the algorithm as the number of dimensions
analyzed was reduced. Figure~\ref{fig:metho-algo-convert} is the output of this
step.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{grayscale.png}
	\caption{Camera capture converted to grayscale}
	\label{fig:metho-algo-convert}
	\centering
\end{figure}

\paragraph{Denoising}
The algorithm denoised the grayscale image using a bilateral filter as
implemented by OpenCV \parencite{opencv-bilateral-filter}. This filter takes the
range of the image into account, rather than just the domain. This resulted in
an image that is smoothed while preserving its edges
\parencite{bilateral-filter}. Figure~\ref{fig:metho-algo-denoise} is the output
of this step.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{denoised.png}
	\caption{Smoothed image with intact edges}
	\centering
	\label{fig:metho-algo-denoise}
\end{figure}

\paragraph{Edge Detection}
The algorithm used a Sobel filter for edge detection. This filter uses a
3$\times$3 kernel that is convolved twice. Once horizontally, and another
vertically to produce a grayscale image of the outlines within the frame. The
kernels used by the Sobel filter \parencite{sobel2014} is shown in
Figure~\ref{fig:metho-algo-key-sobel} and Figure~\ref{fig:metho-algo-sobel} is
the output of this step.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{sobel.png}
	\caption{Output of the Sobel filter}
	\centering
	\label{fig:metho-algo-sobel}
\end{figure}

\paragraph{Thresholding}
The output of the Sobel filter is a grayscale image of 255 values, with each
gray pixel indicating edges within the image. There is a need to reduce the
range of these values to improve the ability of the algorithm to find the
contours of these edges. To do so, the algorithm utilized Otsu's algorithm to
perform automated thresholding. Otsu's algorithm determines a single threshold
that is most optimal for the image \parencite{otsu}. This outputs a
black-and-white image, with only two values. Figure~\ref{fig:metho-algo-thresh}
is the resulting output of this step.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{threshed.png}
	\caption{Threshed image of the edges}
	\label{fig:metho-algo-thresh}
	\centering
\end{figure}

\paragraph{Find Contours}
\label{section:metho-algo-key-contours}
Contours are curves joining all continuous points that have the same color or
intensity \parencite{opencv-contours}. The OpenCV function
\texttt{findContours}, with the \texttt{CHAIN\_APPROX\_SIMPLE} contour
approximation method, was used to find the contours of the outlines of the
object found using the previous step. This approximation method removed
redundant points and returned the least amount of points that describe the
shape. This OpenCV function implements the algorithm of~\cite{contours} in their
paper \citetitle{contours}. Figure~\ref{fig:metho-algo-contours} shows the contours
found in the threshed image.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{find-contours.png}
	\caption{All the contours found in the image}
	\label{fig:metho-algo-contours}
	\centering
\end{figure}

@@

The contour points that described the contour were then sorted clockwise. This
was a necessity since the algorithm of~\cite{contours} does not guarantee that
the largest contour's points were returned in a clockwise arrangement, which is
a requirement for the next step. Figure~\ref{fig:metho-algo-largest-contour}
shows the four contour points found at the edges of the keyboard within the
image.
